Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-08 article/article.tex
Commit 2024-09-08 21:12

In this study, we investigate the characteristics of global horizontal
irradiance enhancement events induced by clouds over Thessaloniki for
the period 1994 -- 2023 using data recorded every one minute. We
identified the cloud enhancement (CE) events by creating an appropriate
cloud-free irradiance reference using a radiative transfer model and
aerosol optical depth data from a collocated Cimel sun photometer and a
Brewer spectrophotometer. We found a trend in CE events of
\(+111\pm 27\,\text{cases}/\text{year}\), and a trend in the
corresponding irradiation of \(+325.2\pm 88\,\text{kJ}/\text{year}\).
The peak of the CE events was observed during May and June. CE events
with duration longer than 10\nobreakspace{}min are very rare
(\(<8\,\%\)), with exceptions lasting over an hour and up to 140
minutes. Finally, we have detected enhancements above the total solar
irradiance at the top of the atmosphere for the same solar zenith angle
of up to \(209\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the
cases below \(40\,\text{W}/\text{m}^{2}\). Most of these extreme events
occur in spring -- early summer, with a secondary peak in autumn.

@@

Following the application of the above discussed methodology to the
entire dataset (\(6\) million of one-minute GHI measurements),
\(2.32\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007 at a SZA of
\(19.9^\circ\) corresponding to OI of \(369.3\,\text{W}/\text{m}^2\) or
\(35.3\,\%\) above the threshold. The strongest CE event of \(\,\%\)
above the cloud-free threshold was observed on 28~October 2016 at a SZA
of \(59.2^\circ\) with a GHI value of \(861.8\,\text{W}/\text{m}^2\) and
a OI of \(298.4\,\text{W}/\text{m}^2\). Both cases are ECE events with
\(131.2\,\text{W}/\text{m}^2\) and \(161.5\,\text{W}/\text{m}^2\) above
the TSI at TOA for the same SZA, respectively. In the following sections
we are discussing the long-term trends and variability of the CE events
as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1994 -- 2023,
together with corresponding linear trends. To account for missing data,
all three quantities have been divided with the fraction of the valid
GHI observations in each month. Statistically significant (at the
\(95\,\%\) confidence level) increasing trends appear for the yearly
number of CE events (\(+111\pm 27\,\text{cases}/\text{year}\)) and the
excess irradiation (\(+325.2 \pm 88\,\text{kJ}/\text{year}\)), while the
the trend in the yearly mean OI is negligible
(\(+0.14\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of no
statistical significance. The average OI for the entire period is
\(+42.6\pm 2.8\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:P-energy-mean}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[\label{fig:P-energy-sum}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1994 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OI and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:P-energy}
\end{figure}

@@

By creating a cloud-free approximation representing the long- and
short-term variability of the expected cloud-free GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1994 -- 2023.
After analyzing the CE cases, we found an increase of
\(+111\pm 26.9\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+325.2\pm 88\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.

>>>>

In this study, we investigate the characteristics of global horizontal
irradiance enhancement events induced by clouds over Thessaloniki for
the period 1994 -- 2023 using data recorded every one minute. We
identified the cloud enhancement (CE) events by creating an appropriate
cloud-free irradiance reference using a radiative transfer model and
aerosol optical depth data from a collocated Cimel sun photometer and a
Brewer spectrophotometer. We found a trend in CE events of
\(+112\pm 35\,\text{cases}/\text{year}\), and a trend in the
corresponding irradiation of \(+329.9\pm 112\,\text{kJ}/\text{year}\).
The peak of the CE events was observed during May and June. CE events
with duration longer than 10\nobreakspace{}min are very rare
(\(<8\,\%\)), with exceptions lasting over an hour and up to 140
minutes. Finally, we have detected enhancements above the total solar
irradiance at the top of the atmosphere for the same solar zenith angle
of up to \(204\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the
cases below \(40\,\text{W}/\text{m}^{2}\). Most of these extreme events
occur in spring -- early summer, with a secondary peak in autumn.

@@

Following the application of the above discussed methodology to the
entire dataset (\(6\) million of one-minute GHI measurements),
\(2.32\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007 at a SZA of
\(19.9^\circ\) corresponding to OI of \(369.3\,\text{W}/\text{m}^2\) or
\(35.3\,\%\) above the threshold. The strongest CE event of \(53\,\%\)
above the cloud-free threshold was observed on 28~October 2016 at a SZA
of \(59.2^\circ\) with a GHI value of \(861.8\,\text{W}/\text{m}^2\) and
a OI of \(298.4\,\text{W}/\text{m}^2\). Both cases are also ECE events
with \(131.2\,\text{W}/\text{m}^2\) and \(161.5\,\text{W}/\text{m}^2\)
above the TSI at TOA for the same SZA, respectively. In the following
sections we are discussing the long-term trends and variability of the
CE events as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1994 -- 2023,
together with corresponding linear trends. To account for missing data,
all three quantities have been divided with the fraction of the valid
GHI observations in each month. Statistically significant (at the
\(95\,\%\) confidence level) increasing trends appear for the yearly
number of CE events (\(+112\pm 35\,\text{cases}/\text{year}\)) and the
excess irradiation (\(+329.9 \pm 112\,\text{kJ}/\text{year}\)), while
the the trend in the yearly mean OI is negligible
(\(+0.11\pm 0.2\,\text{W}/\text{m}^2/\text{year}\)) and of no
statistical significance. The average OI for the entire period is
\(+42.6\pm 2.8\,\text{W}/\text{m}^2\).

!!!! The yearly excess irradiation due to CE events ranges between about
\(6\) and 13 MJ/m\textsuperscript{2}. On average, it is about one fourth
of the highest daily irradiation of \textasciitilde31
MJ/m\textsuperscript{2} under cloud-free conditions recorded at the
summer solstice in Thessaloniki. Therefore, from solar energy
perspective the yearly excess irradiation is only a tiny fraction of the
available solar energy. Although the interannual variability of the mean
OI is rather weak, the variability about the trend lines is quite large
for the number of CE cases and the excess irradiation (panels b and c).
For these two quantities the spread tends to increase with time,
suggesting a significant variability in cloud patterns over the area,
possibly associated to changes in climate.

The interannual variability of the data about the trend lines is quite
large. Furthermore, the spread tends to increase with time (at least for
the quantities of panels b and c), suggesting a significant variability
in cloud patterns over the area, possibly associated to changes in
climate.

@@

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:P-energy-mean}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-2} }\\
            \subfloat[\label{fig:P-energy-sum}]
                {\includegraphics[width=\linewidth]{../images/P-energy-complete-1} }
        }
    \caption{Time series for the period 1994 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OI and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:P-energy}
\end{figure}

@@

By creating a cloud-free approximation representing the long- and
short-term variability of the expected cloud-free GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1994 -- 2023.
After analyzing the CE cases, we found an increase of
\(+112.1\pm 34.9\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+329.9\pm 112\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.
